---
chapter: 2
knit: "bookdown::render_book"
---

# Experimental design {#ch:modesign}

Three experiments associated with the three hypothesis tests will be processed in this paper.

1. Linear relationship vs. no relationship
2. Linear relationship vs. heteroskedastic linear relationship
3. Linear relationship vs. non-linear relationship

Main steps in the experiments are:

1. Simulate data from both the null distribution and the alternative distribution of each hypothesis test
2. Generate corresponding data plots for the simulated data
3. Save data plots as fixed-sized images with labels indicating which distribution they are from 
4. Train a deep learning classifier to recognize different labels
5. Test the model's performance and compute the accuracy
6. Compare the model's performance with human's (exclusive for the first and second experiment)




<!--
**ADD: explanations of choices** 


Three groups of models are included in the two hypothesis tests, linear, nonlinear and heteroskedasticity. So our target is to train the computer to distinguish the residual plots from these three models. 

Although we want to perform this test under multivariate regression setting, we will use only one X variable to generate all different kinds of plots for training. Because the "statistics" we will use is the residual plot, in terms of teaching the computer of reading these plots, one X variable is enough for us to generate different patterns in that plot for convnets to learn. And this also makes the design process much simpler. As for generation of X, we need to consider the distributions of X since it does have an impact on the shape of the scatters. For instance, if X are generated from a uniform distribution, then the plots will look like a square especially when the sample size is large; while more like a circle if X follows normal distribution. In this paper, we are going to fix the distribution of X variables as the same normal distribution $N(0,3)$ for all three groups of data. This way, we treat the X variables as non-random as what we do in practice, so we can compare different relationships of X and Y more easily.

We will set intercept to be zero for all three models, since it has no impact on the patterns of residual plots.

For each group 900 simulated data sets will be generated at first to facilitate deep learning model modification. (The data sets will be augmented later on for training.) 
The sample sizes will be 900 random numbers generated from 20 to 1500. These sample sizes will be kept the same for all the three groups. We choose 20 to 1500, because when the sample size is smaller than 20, say, 10, there is hardly any patterns to see (as shown in figure 2.1). And 1500 is large enough to give a good description about the true relationship within the data, and is light enough to be processed. Since we make the transparency of the points to be 0.01, more points being added to the plot will just make the plot looks darker. Note the upper bound of the sample size is subject to possible changes according to our future study. 


We set image size to be 150 $\times$ 150 pixels, because this size is able to show the plots clearly and also reasonably small so the images can be saved and processed easily.
-->

### Models for relationship

#### Classic linear model - no relationship

This is the null scenario, eg. all Gauss-Markov assumptions are true in the original linear model of X and Y. If the data arises from this model, then the residual plots will not show departures from the assumptions. However, there may still be some undesired patterns formed out of randomness, especially when the sample size is small. Unlike what Simchoni did in his post, we will not use any conventional tests to sort out the "significant" observations. Because we want to know if the deep learning model can distinguish from patterns formed by chance and by nature. Hence, we will simply use the true parent model as the residuals' identification directly for training.

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i, ~~i=1, \dots , n$$

with elements of the model generated by the following processes

- $X\sim N[0,3]$
- $\beta_0 = 0$
- $\beta_1\sim U[0.5, 1]$
- $\varepsilon\sim N(0, 1)$
- $n\sim U[20, 1500]$

<!-- **ADD: explanation of choices, eg error variance=1** -->

After standardization, all residuals will follow standard normal distribution. Therefore, variance of the error term will not make any differences to the residual plots, neither does the correlation coefficient between X and Y. So we will set variance of error term to be one, then randomly select a correlation from 0.5 to 1.  Figure is four examples of residual plots generated under classic linear model.

```{r classic_plot, message=FALSE, cache=TRUE}
library(gridExtra)
classic<-function(n){
  
  beta<-runif(1,0.5,1)
  
  x<-rnorm(n, 0, 3)
  y<-rnorm(n, beta*x, 1)
  
  tibble(x, y)
  model<-lm(y ~ x)
  res<-residuals(model)
  yhat<-predict(model)
  res<-(res-mean(res))/sd(res)
  yhat<-(yhat-mean(yhat))/sd(yhat)
  return(tibble(res, yhat))
}
set.seed(0405)
N <- sample(20:1500, 4, replace=TRUE)
cls <- tibble(N) %>% 
  mutate(res=map(N, classic))
cls <- cls$res

p1 <- ggplot(cls[[1]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) +
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)
p2 <- ggplot(cls[[2]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) + 
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)
p3 <- ggplot(cls[[3]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) + 
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)
p4 <- ggplot(cls[[4]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) + 
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(1,1)) 

```

#### Classic linear model - important variable left out

This level of the experiment is important to be able to do the comparison with the human subject study. 

Sometimes people want to test if certain variables (X2) are omitted from the regression, to do this, we can make scatter plot of residuals on the suspected omitted variable (X2) and test if there is significant linear relationship between them.

In order to do the comparison, we first use the same specification with the experiment two in @MM13 to generate data for training and testing purpose.

$$Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \varepsilon_i, ~~i=1, \dots , n$$
with elements of the model generated by the following processes

- $X_1, X_2\sim N[0,3]$  
- $\beta_0 = 0$  
- $\beta_1\sim U[0.5, 1]$  
- $\varepsilon\sim N(0, 5) \ or\ N(0,12)$  
- $n\sim [100, 300]$  

In this case, $X_2$ will be the omitted variable, and $\beta_2$ will be generated uniformly from -10 to -0.1 or from 0.1 to 10 in order to cover the range of which in experiment two. To simplify the comparison, we set the sample size to be 100 to 300 in this case. Since the human experiment only used sample size of 100 or 300. The same for standard deviation of error term, we set it to be 5 or 12 to be consistent with the human experiment. Below four plots are examples generated from this specification.

```{r linear}
library(gridExtra)
linear<-function(n){
   beta <- sample(c((-10:-0.1),(0.1:10)), 1)
  
  sigma <- sample(1:12 , 1)
  x<-rnorm(n, 0, 1)
  y<-rnorm(n, beta*x, sigma)
  
  res <- y
  yhat <- x
  tibble(res, yhat) %>% 
  ggplot(aes(x = yhat, y=res)) + 
    geom_point(alpha = 0.4) +
    theme(aspect.ratio = 1)
}
set.seed(0420)
n1 <- sample(c(100, 300), 1)
n2 <- sample(c(100, 300), 1)
n3 <- sample(c(100, 300), 1)
n4 <- sample(c(100, 300), 1)

p1 <- linear(n1)
p2 <- linear(n2)
p3 <- linear(n3)
p4 <- linear(n4)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(1,1)) 
```


#### Linear Model with Heteroskedasticity {#sec:heter}

<!-- **CHANGE THIS SECTION TO BE EXPLICIT LIKE THE CLASSIC MODEL** -->

This model is designed to represent one of the alternative hypothesis, eg. when the constant variance assumption is violated holding all other conditions unchanged.

Hence, the only difference needs to be made for this model is to add a non-zero relationship between the variance of Y and X. It is impossible to list out all types of heteroskedasticity by a simple function form. Here I am trying to present four kinds of relationship between variance of Y and X using a second order polynomial. With randomly generated parameters, the variance of Y could be monotonically increasing/decreasing with X, or has a turning point in the middle.

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i, ~~i=1, \dots , n$$

with elements of the model generated by the following processes

- $X\sim N[0,\ 3]$
- $\beta_0 = 0$
- $\beta_1\sim U[0.5,\ 1]$
- $\varepsilon_i\sim N(0,\ ax_i^2+bx_i+c+v_i)$
- $v_i \sim N(0,\ 0.25)$
- $n\sim U[20,\ 1500]$

Choices of the distribution to generate a, b, c and v is mainly an empirical work. With freedom of choosing a and b, all four types of relationships can be realized. Primarily, we want the residual plots to show variations among different data sets; on the other hand, we need to take a bit control on the choices of these parameters, prevent the final variance from being too large so the linear relationship between X and Y will not disappear. Figure below is four examples of residual plots generated under this model.

```{r heter_plot, message=FALSE, cache=TRUE}
heter<-function(n){
  x<-rnorm(n, 0, 3)
  beta<-runif(1,0.5,1)
  
  a <- runif(1,0,2)*rbinom(1,1,0.5)
  b <- runif(1,-4,4)
  c <- rnorm(1,0,2)
  
  variance <- 0.25*(a*x^2+b*x+c)+rnorm(n,0,0.25)
  
  index <- sample(0:1,1)
  if (index==1) {
    variance <- -variance
  }
  
  min <- min(variance)
  
  if (min<0){
    variance <- variance-min
  }
  
  max <- max(variance)
  while (max>5) {
    variance <- 0.8* variance
    max <- max(variance)
  }
  
  y<-rnorm(n, beta*x, variance)
  
  tibble(x, y)
  model<-lm(y ~ x)
  res<-residuals(model)
  yhat<-predict(model)
  res<-(res-mean(res))/sd(res)
  yhat<-(yhat-mean(yhat))/sd(yhat)
  return(tibble(res, yhat))
}
set.seed(0003)
N <- sample(20:1500, 4, replace=TRUE)
het <- tibble(N) %>% 
  mutate(res=map(N, heter))
het <- het$res

p1 <-   ggplot(het[[1]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)
p2 <-   ggplot(het[[2]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)
p3 <-   ggplot(het[[3]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)
p4 <-   ggplot(het[[4]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)

grid.arrange(p1,p2,p3,p4, nrow=2) 

```

<!--
Since we do not have existing data from human experiments, to provide a reference to evaluate how computer perform in detecing heteroskedasticity issues, a special case of the White test will be used. Each image in test set will be evaluated using this test and the associated accuracy will be compared to computers' accuracy on the same set.

-->

#### Non-linear model {#sec:poly}

This data set will be representative of non-linear relationship between X and Y, keep all other factors being the same with the classic linear model.

Obviously we cannot write down all different non-linear relationships between X and Y. Luckily, we do not need to actually do that. Because our purpose is for the model to tell if there is any kind of non-linear patterns, but not to specify which type it is. As long as the relationship between residual plots is curved, we do not worry about if it is from a logarithm function or squared root. Therefore all we want is for the model to "see" varied patterns, we just need to find a way to show different curvature. Hence, at this stage, we only include a polynomial of X to the order of five with each parameter being randomly generated. I chose five because normally the order of the non-linear relationship of data from Economics world do not exceed four. What's more, we find this design is good enough to vary the pattern of our plots. Also, since order four and five are not as common as lower orders, we attach a bernoulli term with 0.5 success rate to them respectively. 

So the Y will be a normal distribution with mean
$\beta_1x+\beta_2x^2+\beta_3x^3+\beta_4x^4+\beta_5x^5$, and variance $\beta_1^2+\beta_2^2+\beta_3^2+2\beta_4^2+2\beta_5^2$. We adjust the variance to be increase for higher orders so the data could has more variation. Below are four examples generated from this model.

```{r poly_plot, message=FALSE, cache=TRUE}
nonpoly<-function(n){
  
  p1<-rnorm(1,0,2)
  p2<-rnorm(1,0,2)
  p3<-rnorm(1,0,2)*rbinom(1,1,0.5)
  p4<-rnorm(1,0,2)*rbinom(1,1,0.3)
  p5<-rnorm(1,0,1)*rbinom(1,1,0.2)
  
  x<-runif(n,-3,3)
  y<-rnorm(n, p1*x+p2*x^2+p3*x^3+p4*x^4+p5*x^5, p1^2+2*p2^2+2*p3^2+2*p4^2+2*p5^2)
  
  tibble(x, y)
  model<-lm(y ~ x)
  res<-residuals(model)
  yhat<-predict(model)
  res<-(res-mean(res))/sd(res)
  yhat<-(yhat-mean(yhat))/sd(yhat)
  return(tibble(res, yhat))
}

set.seed(0002)
N <- sample(20:800, 4, replace=TRUE)
pl <- tibble(N) %>% 
  mutate(res=map(N, nonpoly))
pl <- pl$res

p1 <-   ggplot(pl[[1]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)
p2 <-   ggplot(pl[[2]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)
p3 <-   ggplot(pl[[3]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)
p4 <-   ggplot(pl[[4]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(0.6,0.6)) 
```




## Extract residuals and fitted values

We will design three functions to generate three different types of data. Within each function, we will return the standardized residuals and predicted value of Y from the fitted linear model. After running each funtion N times (N=900 for the early stage), we will get three lists each containing N sets of residuals and predicted Y.

## Scale data into fixed image size

We will use ggplot2 package in R to generate residual plots and use ggsave funtion to resize and save all plots to our local drive. So all plots will be a squared image with same width and height. This helps us to see better the patterns in the scatter plot and also is easier for the deep learning model to process the image data.

## Train the deep learning classifier

We will use Keras package in R to implement the convolutional neural network. For now, given the computation capibility, we have simulated 900 images of each group as our training set; 200 images of each group as the validation set; and 300 images of each group as the unseen test set. These sets will be enlarged later on to improve the model's ability on unseen data.

## Assess the accuracy

After 30 iterations, the accuracy of the training set has reached 99.25%, 93.3% for validation set, and 94% for test set. This performance is consistent with our expectation of the deep learning model. 
























