---
chapter: 2
knit: "bookdown::render_book"
---

# Experimental design {#ch:modesign}

Three experiments associated with the three hypothesis tests will be processed in this paper.

1. Linear relationship vs. no relationship
2. Linear relationship vs. heteroskedastic linear relationship
3. Linear relationship vs. non-linear relationship

Main steps in the experiments are:

1. Simulate data from both the null distribution and the alternative distribution of each hypothesis test
2. Generate corresponding data plots for the simulated data
3. Save data plots as fixed-sized images with labels indicating which distribution they are from 
4. Train a deep learning classifier to recognize different labels
5. Test the model's performance and compute the accuracy
6. Compare the model's performance with human's (exclusive for the first and second experiment)

## Data simulation and model design

### Classic linear model

This is the model implied in the alternative hypothesis of the first experiment in this paper, and it is similar to what @SIM18 did in his blog. After the deep learning model is trained on data generated in this section, it will be tested on the 70 data plots from the second experiment in @MM13. This model's accuracy will then be compared to the human's performance.

The model is designed as:
$$Y_i = \beta_0 + \beta_1 X_{i}  + \varepsilon_i, ~~i=1, \dots , n$$
with elements of the model generated by the following processes

- $X \sim N[0,\ 1]$  
Distributions of X has impact on the shape of the scatters. For instance, if X is generated from a uniform distribution, then the plots will look like a square especially when the sample size is large; while more like a circle if X follows normal distribution. In this experiment, we will set it to be normal to match with the design in the second experiment of @MM13. 

- $\beta_0 = 0$  
Intercept is set to be zero, because it will not change the data plots since all data will be standardized before being plotted. Any constant intercept has same effect on the data plots.

- $\beta_1\sim U[-10, -0.1] \bigcup [0.1, 10]$  
$\beta_1$ is designed to be uniformly generated from -10 to 10 (excluding -0.1 to 0.1) to cover the range used in the second experiment in @MM13.

- $\varepsilon\sim N(0, \sigma^2) \ where\ \sigma \sim U(1,12)$  
$\varepsilon$ is designed to be uniformly generated from 1 to 12 in order to cover the range used in the second experiment in @MM13.

- $n=100, 300$  
Number of observations are the same with the the second experiment in @MM13.

```{r linear, echo=FALSE, fig.cap="Four examples of data plots generated from the classic linear model"}
library(gridExtra)
linear<-function(n){
   beta <- sample(c((-10:-0.1),(0.1:10)), 1)
  
  sigma <- sample(1:12 , 1)
  x<-rnorm(n, 0, 1)
  y<-rnorm(n, beta*x, sigma)
  
  tibble(x, y) %>% 
  ggplot(aes(x = x, y = y)) + 
    geom_point(alpha = 0.4) +
    theme(aspect.ratio = 1)
}
set.seed(0420)
n1 <- sample(c(100, 300), 1)
n2 <- sample(c(100, 300), 1)
n3 <- sample(c(100, 300), 1)
n4 <- sample(c(100, 300), 1)

p1 <- linear(n1)
p2 <- linear(n2)
p3 <- linear(n3)
p4 <- linear(n4)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(1,1)) 
```

### Linear Model with Heteroskedasticity

This is the model implied in the alternative hypothesis of the second experiment in this paper, where the constant variance assumption of the linear model is violated while all other conditions are met. 
By the definition given in @IE17, "The homoskedasticity states that the variance of the unobserved error, u, conditional on the explanatory variables, is constant. Homoskedasticity fails whenver the variance of the unobserved factors changes across different segments of the population, where the segments are determined by the different values of the explanatory variables."
There are countless types of heteroskedasticity since the "change of the variance" could be related to "the explanatory variables" in various ways. It is not feasible to list out all kinds of heteroskedasticity by a single function. For simplicity, we will focus on one example of them, a linear correlation between the explanatory variable X and the standard deviation of the error term. The conclusion from this experiment though can be generalized to more complicated cases.  
A new human experiment will be rendered to produce the human's performance on detecting heteroskedasticity by reading residual plots. After the deep learning model is trained on the data generated from this section, it will be tested on the same data set that is used for the human experiment. And the accuracy of the model will be compared to the human's performance from this new human experiment. 

The model structure is the same with the classic linear model:
$$Y_i = \beta_0 + \beta_1 X_{i}  + \varepsilon_i, ~~i=1, \dots , n$$
with elements of the model generated by the following processes

- $X\sim U[-1,\ 1]$
- $\beta_0 = 0$
- $\beta_1\sim U[0.5,\ 1]$
- $\varepsilon\sim N(0,\ ax+v)$
- $a\sim U(0.05,\ 5)$
- $v \sim N(0,\ 1)$
- $n\sim U[20,\ 1500]$





Choices of the distribution to generate a, b, c and v is mainly an empirical work. With freedom of choosing a and b, all four types of relationships can be realized. Primarily, we want the residual plots to show variations among different data sets; on the other hand, we need to take a bit control on the choices of these parameters, prevent the final variance from being too large so the linear relationship between X and Y will not disappear. Figure below is four examples of residual plots generated under this model.

To provide a reference to evaluate how computer perform in detecing heteroskedasticity issues, a special case of the White test will be used. Each image in test set will be evaluated using this test and the associated accuracy will be compared to computers' accuracy on the same set.

```{r heter_plot, message=FALSE, cache=TRUE}
heter<-function(n){
  x<-rnorm(n, 0, 3)
  beta<-runif(1,0.5,1)
  
  a <- runif(1,0,2)*rbinom(1,1,0.5)
  b <- runif(1,-4,4)
  c <- rnorm(1,0,2)
  
  variance <- 0.25*(a*x^2+b*x+c)+rnorm(n,0,0.25)
  
  index <- sample(0:1,1)
  if (index==1) {
    variance <- -variance
  }
  
  min <- min(variance)
  
  if (min<0){
    variance <- variance-min
  }
  
  max <- max(variance)
  while (max>5) {
    variance <- 0.8* variance
    max <- max(variance)
  }
  
  y<-rnorm(n, beta*x, variance)
  
  tibble(x, y)
  model<-lm(y ~ x)
  res<-residuals(model)
  yhat<-predict(model)
  res<-(res-mean(res))/sd(res)
  yhat<-(yhat-mean(yhat))/sd(yhat)
  return(tibble(res, yhat))
}
set.seed(0003)
N <- sample(20:1500, 4, replace=TRUE)
het <- tibble(N) %>% 
  mutate(res=map(N, heter))
het <- het$res

p1 <-   ggplot(het[[1]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)
p2 <-   ggplot(het[[2]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)
p3 <-   ggplot(het[[3]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)
p4 <-   ggplot(het[[4]], aes(x=yhat, y=res)) +geom_point(alpha=0.4) +
  theme(aspect.ratio = 1)

grid.arrange(p1,p2,p3,p4, nrow=2) 

```



<!--
**ADD: explanations of choices** 


For each group 900 simulated data sets will be generated at first to facilitate deep learning model modification. (The data sets will be augmented later on for training.) 
The sample sizes will be 900 random numbers generated from 20 to 1500. These sample sizes will be kept the same for all the three groups. We choose 20 to 1500, because when the sample size is smaller than 20, say, 10, there is hardly any patterns to see (as shown in figure 2.1). And 1500 is large enough to give a good description about the true relationship within the data, and is light enough to be processed. Since we make the transparency of the points to be 0.01, more points being added to the plot will just make the plot looks darker. Note the upper bound of the sample size is subject to possible changes according to our future study. 


We set image size to be 150 $\times$ 150 pixels, because this size is able to show the plots clearly and also reasonably small so the images can be saved and processed easily.
-->

### no relationship

This hypothesis test is very similar to @SIM18. 

In order to do the comparison, we first use the same specification with the experiment two in @MM13 to generate data for training and testing purpose.




This is the null scenario, eg. all Gauss-Markov assumptions are true in the original linear model of X and Y. If the data arises from this model, then the residual plots will not show departures from the assumptions. However, there may still be some undesired patterns formed out of randomness, especially when the sample size is small. Unlike what Simchoni did in his post, we will not use any conventional tests to sort out the "significant" observations. Because we want to know if the deep learning model can distinguish from patterns formed by chance and by nature. Hence, we will simply use the true parent model as the residuals' identification directly for training.

$$Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i, ~~i=1, \dots , n$$

with elements of the model generated by the following processes

- $X\sim N[0,3]$
- $\beta_0 = 0$
- $\beta_1\sim U[0.5, 1]$
- $\varepsilon\sim N(0, 1)$
- $n\sim U[20, 1500]$

<!-- **ADD: explanation of choices, eg error variance=1** -->

After standardization, all residuals will follow standard normal distribution. Therefore, variance of the error term will not make any differences to the residual plots, neither does the correlation coefficient between X and Y. So we will set variance of error term to be one, then randomly select a correlation from 0.5 to 1.  Figure is four examples of residual plots generated under classic linear model.

```{r classic_plot, message=FALSE, cache=TRUE}
library(gridExtra)
classic<-function(n){
  
  beta<-runif(1,0.5,1)
  
  x<-rnorm(n, 0, 3)
  y<-rnorm(n, beta*x, 1)
  
  tibble(x, y)
  model<-lm(y ~ x)
  res<-residuals(model)
  yhat<-predict(model)
  res<-(res-mean(res))/sd(res)
  yhat<-(yhat-mean(yhat))/sd(yhat)
  return(tibble(res, yhat))
}
set.seed(0405)
N <- sample(20:1500, 4, replace=TRUE)
cls <- tibble(N) %>% 
  mutate(res=map(N, classic))
cls <- cls$res

p1 <- ggplot(cls[[1]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) +
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)
p2 <- ggplot(cls[[2]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) + 
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)
p3 <- ggplot(cls[[3]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) + 
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)
p4 <- ggplot(cls[[4]], aes(x=yhat, y=res)) + 
  geom_point(alpha = 0.4) + 
  xlab("Fitted values") + ylab("Residuals") + theme(aspect.ratio = 1)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(1,1)) 

```









### Non-linear model {#sec:poly}

This data set will be representative of non-linear relationship between X and Y, keep all other factors being the same with the classic linear model.

Obviously we cannot write down all different non-linear relationships between X and Y. Luckily, we do not need to actually do that. Because our purpose is for the model to tell if there is any kind of non-linear patterns, but not to specify which type it is. As long as the relationship between residual plots is curved, we do not worry about if it is from a logarithm function or squared root. Therefore all we want is for the model to "see" varied patterns, we just need to find a way to show different curvature. Hence, at this stage, we only include a polynomial of X to the order of five with each parameter being randomly generated. I chose five because normally the order of the non-linear relationship of data from Economics world do not exceed four. What's more, we find this design is good enough to vary the pattern of our plots. Also, since order four and five are not as common as lower orders, we attach a bernoulli term with 0.5 success rate to them respectively. 

So the Y will be a normal distribution with mean
$\beta_1x+\beta_2x^2+\beta_3x^3+\beta_4x^4+\beta_5x^5$, and variance $\beta_1^2+\beta_2^2+\beta_3^2+2\beta_4^2+2\beta_5^2$. We adjust the variance to be increase for higher orders so the data could has more variation. Below are four examples generated from this model.

```{r poly_plot, message=FALSE, cache=TRUE}
nonpoly<-function(n){
  
  p1<-rnorm(1,0,2)
  p2<-rnorm(1,0,2)
  p3<-rnorm(1,0,2)*rbinom(1,1,0.5)
  p4<-rnorm(1,0,2)*rbinom(1,1,0.3)
  p5<-rnorm(1,0,1)*rbinom(1,1,0.2)
  
  x<-runif(n,-3,3)
  y<-rnorm(n, p1*x+p2*x^2+p3*x^3+p4*x^4+p5*x^5, p1^2+2*p2^2+2*p3^2+2*p4^2+2*p5^2)
  
  tibble(x, y)
  model<-lm(y ~ x)
  res<-residuals(model)
  yhat<-predict(model)
  res<-(res-mean(res))/sd(res)
  yhat<-(yhat-mean(yhat))/sd(yhat)
  return(tibble(res, yhat))
}

set.seed(0002)
N <- sample(20:800, 4, replace=TRUE)
pl <- tibble(N) %>% 
  mutate(res=map(N, nonpoly))
pl <- pl$res

p1 <-   ggplot(pl[[1]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)
p2 <-   ggplot(pl[[2]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)
p3 <-   ggplot(pl[[3]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)
p4 <-   ggplot(pl[[4]], aes(x=yhat, y=res)) +geom_point(alpha = 0.4) + theme(aspect.ratio = 1)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(0.6,0.6)) 
```

## Generate corresponding data plots for the simulated data

We will design three functions to generate three different types of data. Within each function, we will return the standardized residuals and predicted value of Y from the fitted linear model. After running each funtion N times (N=900 for the early stage), we will get three lists each containing N sets of residuals and predicted Y.

## Save data plots as fixed-sized images with labels indicating which distribution they are from

We will use ggplot2 package in R to generate residual plots and use ggsave funtion to resize and save all plots to our local drive. So all plots will be a squared image with same width and height. This helps us to see better the patterns in the scatter plot and also is easier for the deep learning model to process the image data.

##Train a deep learning classifier to recognize different labels

We will use Keras package in R to implement the convolutional neural network. For now, given the computation capibility, we have simulated 900 images of each group as our training set; 200 images of each group as the validation set; and 300 images of each group as the unseen test set. These sets will be enlarged later on to improve the model's ability on unseen data.

## Test the model's performance and compute the accuracy

After 30 iterations, the accuracy of the training set has reached 99.25%, 93.3% for validation set, and 94% for test set. This performance is consistent with our expectation of the deep learning model. 

## Compare the model's performance with human's (exclusive for the first and second experiment)






















