---
chapter: 2
knit: "bookdown::render_book"
---

# Comparing computer performance against database of human evaluation

## Turk study explanation

A large database of results from a human subjects test conducted to validate the lineup protocol relative to a classical tests is going to be used for this part of the work. This data was collected as part of the work presented in @MM13. Experiment 2 examined the performance of humans in recognising linear association between two variables, in direct comparison to conducting a $t$-test of $H_o: \beta_k=0$ vs $H_a: \beta_k\neq 0$ assessing the importance of including variable $k$ in the linear model. An example lineup is shown in Figure \ref{expt2}. For this lineup, 63 of the 65 people who examined it selected the data plot (position 20) from the null plots. There is clear evidence that the data displayed in plot 20 is not from $H_o: \beta_k=0$. 

\begin{figure*}[h]
\centerline{\includegraphics[width=15cm]{figures/plot_turk2_300_350_12_3.png}}
\caption{One of 70 lineups used in experiment 2 Majumder et al (2012). Of the 65 people who examined the lineup,  63 selected the data plot, which is in position 20.}
\label{expt2}
\end{figure*}

This experiment utilised 70 lineups of size 20 plot, with varying degrees of departure from the $H_o: \beta_k=0$. There were 351 evaluations by human subjects. These results will be used for comparison with the deep learning model. 

The trained deep learning model will be applied to the data from this experiment. The model will be asked to classify the 70 data plots. We will calculate how frequently the data plot is selected as not a null plot, and compare this to the frequencies obtained by human evaluation. 

## Linear relationship simulation

The design for this model is similar to what @SIM18 did in his blog but is tailored to perform the comparison with human performance later on.

The model is designed as:
$$Y_i = \beta_0 + \beta_1 X_{i}  + \varepsilon_i, ~~i=1, \dots , n$$
with elements of the model generated by the following processes

- $X \sim N[0,\ 1]$  
Distributions of X has impact on the shape of the scatters. For instance, if X is generated from a uniform distribution, then the plots will look like a square especially when the sample size is large; while more like a circle if X follows normal distribution. In this experiment, we will set it to be normal to match with the design in the second experiment of @MM13. 

- $\beta_0 = 0$  
Intercept is set to be zero, because it will not change the data plots since all data will be standardized before being plotted. Any constant intercept has same effect on the data plots.

- $\beta_1\sim U[-10, -0.1] \bigcup [0.1, 10]$  
$\beta_1$ is designed to be uniformly generated from -10 to 10 (excluding -0.1 to 0.1) to cover the range used in the second experiment in @MM13.

- $\varepsilon\sim N(0, \sigma^2) \ where\ \sigma \sim U(1,12)$  
$\varepsilon$ is designed to be uniformly generated from 1 to 12 in order to cover the range used in the second experiment in @MM13.

- $n=100, 300$  
Number of observations are the same with the the second experiment in @MM13.

```{r linear, echo=FALSE, fig.cap="Four examples of data plots generated from the classic linear model"}
library(gridExtra)
linear<-function(n){
   beta <- sample(c((-10:-0.1),(0.1:10)), 1)
  
  sigma <- sample(1:12 , 1)
  x<-rnorm(n, 0, 1)
  y<-rnorm(n, beta*x, sigma)
  
  tibble(x, y) %>% 
  ggplot(aes(x = x, y = y)) + 
    geom_point(alpha = 0.4) +
    theme(aspect.ratio = 1)
}
set.seed(0420)
n1 <- sample(c(100, 300), 1)
n2 <- sample(c(100, 300), 1)
n3 <- sample(c(100, 300), 1)
n4 <- sample(c(100, 300), 1)

p1 <- linear(n1)
p2 <- linear(n2)
p3 <- linear(n3)
p4 <- linear(n4)

grid.arrange(p1,p2,p3,p4,nrow=2, heights=c(1,1)) 
```

## Null plots simulation

This is the null scenario in the first experiment, eg. the two variables under tested are independent of each other. If the data arises from this situation, then the data plots will not show any systematic patterns theoretically. 

The model is designed as:
$$Y_i = \beta_0 + \beta_1 X_{i}  + \varepsilon_i, ~~i=1, \dots , n$$
with elements of the model generated by the following processes

- $X \sim N[0,\ 1]$  
- $\beta_0 = 0$  
- $\beta_1 = 0$  
- $\varepsilon\sim N(0, \sigma^2) \ where\ \sigma \sim U(1,12)$  
- $n=100, 300$  

The specification of this model is the same with the linear relationship, the only difference is the correlation coefficient equals to zero in this case.

```{r norela, echo=F, fig.cap="Four examples of data plots generated with two independent variables"}
norela<-function(i){
  n <- sample(c(100, 300), 1)
  beta <- 0
  sigma <- sample(1:12 , 1)
  x<-rnorm(n, 0, 1)
  y<-rnorm(n, beta*x, sigma)
  pic <- tibble(x, y) %>% 
    ggplot(aes(x = x, y=y)) + 
    geom_point(alpha = 0.4) +
    theme(aspect.ratio = 1)
}
set.seed(0518)
p1 <- norela(1)
p2 <- norela(2)
p3 <- norela(3)
p4 <- norela(4)
grid.arrange(p1,p2,p3,p4, nrow=2)

```

## Deep learning model setup

The plot used for training and testing in this section is the scatter plot between the dependent variable Y and the independent variable X. "ggplot2" package in R is used to generate the plots and to save all plots to our local drive. All plots will be resized to be squared images with same width and height. This helps us to see better the patterns in the scatter plot and also is easier for the deep learning model to process. As for the labels given the each image, we will use the true population as the samples' identification directly. It is true that there will be some undesired patterns formed out of randomness, especially when the sample size is small. Unlike what Simchoni did in his post, no conventional tests will be used to sort out the "significant" observations. Because the answer to the question that if the deep learning model can distinguish from patterns formed by chance and by nature is also interested. 

All convolutional neural network related work will be done by Keras package in R.

```{r dl_model, echo=FALSE, fig.cap="The deep learning model structure used in this thesis"}
model_loaded <- load_model_hdf5("data/weights.06-0.22.hdf5")
model_loaded
```

180,000 data sets are generated for each of the two groups in the first experiment where 100,000 of them is used for training; 40,000 of them is used for validation and 40,000 for testing. 10 epoches (1 epoch = 1 iteration over all samples) will be done for training. The model specification given by each epoch will be saved, the one gives the highest test power is chosen to represent the computers. 

\begin{figure*}[h]
\centerline{\includegraphics[width=15cm]{figures/new_plot.png}}
\caption{Plot of the training history of the deep learning model for the first experiment.}
\label{expt2}
\end{figure*}

From the plot of the traning history, we can see the fourth specification gives the overall best performance with lowest loss and highest accuracy on validation set. Its accuracy on the unseen test set is also the best among other models. After the fourth checkpoint, the model starts to overfit.

```{r checkpoints, echo=F}
checkpoints <- c("4 epoch", "6 epoch", "10 epoch", "T-test")
powers <- c(0.8790, 0.8910, 0.8826, 0.9003)
alpha_1 <- c(0.9757, 0.9569, 0.9687, 0.9487) 
accuracy <- c(0.9273, 0.9239, 0.9257, 0.9245)
tibble(checkpoints, powers, alpha_1, accuracy) %>% kable(col.names = c("Tests", "Power", "1-Alpha", "Overall accuracy"), caption = "Performance of three checkpoints of the deep learning model on the unseen test set, where the power means the accuracy on the linear ones, (1-alpha) is the accuracy on the no-relationship ones. All numbers have been rounded to four decimal places.")
```

However, since the majority of the data plots in Turk's experiment have been generated with linear relationship, we will choose the sixth model here because it has the highest test power which provides the deep learning model a better chance. What's more, the $\alpha$ of the sixth model, which is the probability of Type I error, is the closest to 0.05. We can see the sixth model's performance is quite close to the conventional T-test.

## Comparing results

```{r linearresults, echo=F}
tests <- c("human", "computer", "t-test")
correct <- c(47, 44, 43)
accuracy <- c(0.6714, 0.6286, 0.6143)
tibble(tests, correct, accuracy) %>% kable(col.names = c("Tests", "No. of correct", "Accuracy"), caption = "Accuracy of testing the 70 data plots evaluated by human computer and the conventional t-test.")
```

The conclusion of human evaluation is calculated using the p-value calculation introduced in section 2 of @MM13, the null hypothesis is rejected if the calculated p-value is smaller than 0.05. 

The $\alpha$ of the conventional t-test we used in this comparison is 0.05. 

From this table, we can see human evaluation achieves the best accuracy on the Turk's experiment data set. Deep learning model and the conventional t-test perform similarly. 
